---
title: "Foundations of Estimation in Biomedical Studies"
author: "<br><br><span style='font-size:25px;'><strong>Lesley Chapman Hannah, Ph.D., M.S.</strong></span><br>College of Graduate Studies<br>Northeast Ohio Medical University"

format: pptx

execute:
  echo: true
  warning: false
  message: false
---

```{r, echo=FALSE}
library(tidyverse)
library(broom)
library(scales)
set.seed(104)

biomarker_df <- tibble(
  id = 1:180,
  treatment = rep(c("Placebo", "Low dose", "High dose"), each = 60),
  ifng = c(
    rnorm(60, mean = 11.5, sd = 4.2),
    rnorm(60, mean = 15.3, sd = 4.8),
    rnorm(60, mean = 19.8, sd = 5.1)
  ),
  response = c(
    rbinom(60, 1, 0.28),
    rbinom(60, 1, 0.45),
    rbinom(60, 1, 0.62)
  ),
  infections = c(
    rpois(60, lambda = 1.4),
    rpois(60, lambda = 1.1),
    rpois(60, lambda = 0.8)
  ),
  age = round(rnorm(180, mean = 61, sd = 10)),
  bmi = round(rnorm(180, mean = 29, sd = 4.5), 1)
)
```

### Lecture Overview

-   define estimation in biomedical and clinical data analysis
-   provide context for the intersection between estimation, descriptive statistics and formal inference
-   explain how sampling variability motivates interval estimation

------------------------------------------------------------------------

### Estimation in Biomedical Data Analysis

::: {style="font-size:0.5em"}
> Estimation uses the observed sample to quantify unknown population values, expressing both the best-supported value and the uncertainty introduced by sampling variability

Biomedical and clinical research is conducted on samples instead of data from entire populations

Each dataset represents:

-   subset of patients
-   observed under specific study conditions
-   measured at a defined point in time

Within the context of biomedical research the purpose of analyzing a sample is to learn about \[select\]:

-   broader patient population
-   underlying biological processes
-   treatment effects
:::

------------------------------------------------------------------------

### Estimation Definition

::: {style="font-size:0.68em"}
> Estimation: process of using sample data to produce quantitative approximations of unknown population or biological parameters that inform biomedical and clinical understanding

In practice, estimation allows us to:

-   pproximate a true mean biomarker level
-   approximate a true probability of response
-   approximate a true rate of clinical events
:::

------------------------------------------------------------------------

### Whats being estimated?

::: {style="font-size:0.68em"}
estimation targets defined population quantities, including:

-   `probabilities`: proportion of patients with a mutation
-   `rates`: mutation counts per genomic region
-   `means`: average biomarker levels
-   `effects`: associations between predictors and outcomes

quantities not directly observed $\rightarrow$ estimated from sample data

estimation goal: produce values that reflect these underlying quantities as accurately as possible given the data available
:::

------------------------------------------------------------------------

### Review: Population Parameters and Sample Estimators

::: {style="font-size:0.58em"}
-   `population`: the full set of individuals or biological units of interest, such as all patients with a specific disease

-   `sample`: the subset of individuals that are actually observed in a study

-   to estimate an unknown population parameters we use sample data as a basis for estimating the unknown population parameters because the full population parameter cannot be observed

since we usually can’t observe the whole population, we rely on sample data to approximate unknown parameters - numerical quantity that describes a population or biological process:

$$
\mu \approx \bar{x}, \qquad p \approx \hat{p}, \qquad \lambda \approx \hat{\lambda}
$$ - population mean μ is estimated by the sample mean $\bar{x}$

-   population proportion p is estimated by the sample proportion $\hat{p}$

-   population rate 𝜆is estimated by an observed sample rate $\hat{\lambda}$
:::

------------------------------------------------------------------------

### Estimation Techniques Definition

> estimation techniques are methods for using sample data to produce quantitative estimates of unknown population or biological parameters that inform clinical and biomedical decisions

Statistical estimation techniques are the primary tools used to quantify biological signals from observed data and translate them into population-level quantities

------------------------------------------------------------------------

### Descriptive Statistics, Estimation, and Inference

::: {style="font-size:0.68em"}
-   descriptive statistics summarize the observed sample using quantities such as the mean, median, standard deviation, proportion, and range
    -   what are the main features of this sample?
-   estimation uses those summaries as tools for approximating **unknown population quantities** such as a true mean biomarker level, a true treatment response probability, or a true event rate
    -   what does this sample suggest about the population or biological process that generated it?
-   inference estimates:
    -   process of quantifying uncertainty and making decisions based on data, often under repeated sampling
    -   variable quantity, how variable the estimate would be if we repeated the study many times
    -   confidence intervals and hypothesis tests are tools that formalize this variability.
:::

------------------------------------------------------------------------

### Estimation and Sampling

::: {style="font-size:0.68em"}
estimates are computed from sampled data rather than the full population.

datasets represents:

-   subset of individuals

-   observed under specific study conditions

-   measured at a defined point in time

-   value of an estimate depends on which observations are included in the sample

-   different sample from the same population may produce a different estimate
:::

------------------------------------------------------------------------

### Sampling Variability

::: {style="font-size:0.68em"}
-   `sampling variability`: variation in a statistic across repeated random samples from the same population, reflecting that each sample provides only partial information about the population parameter it is used to estimate
    -   estimates vary because each sample captures a different subset of the population
    -   sampling variability key concept for uncertainty quantification
    -   defines how estimates behave when data are collected from patient populations
    -   confidence intervals, standard errors, and other measures of precision are derived from the distribution of estimates across repeated samples
-   estimation involves using sampled data to characterize an unknown population quantity
    -   estimate depends on which observations are included and will vary across different possible samples
:::

------------------------------------------------------------------------

### Research Example

::: {style="font-size:0.68em"}
**Background**:

-   consider a clinical study of metabolic health in adults receiving routine outpatient care
-   study aim : learn about the broader patient population, not only the specific individuals who were enrolled in one dataset

Questions of interest include estimating the following targets linekd to an unknonw population quantity:

-   What is the average fasting glucose level in the target population?
-   What proportion of patients meet criteria for metabolic syndrome?
-   What is the expected number of follow-up visits per patient?
-   How are these outcomes associated with age and body mass index?
:::

------------------------------------------------------------------------

## Research Objectives

::: {style="font-size:0.68em"}
-   identify the population quantity being targeted
-   distinguish a parameter from its sample-based estimate
-   explain why estimates vary across samples
-   quantify the variability of an estimate
-   interpret confidence intervals as estimation tools
-   connect simple estimation to model-based estimation
:::

------------------------------------------------------------------------

### Example Dataset: Simulated Population

::: {style="font-size:0.68em"}
Simulated Dataset: true population quantities with sample-based estimates

The population contains:

-   age
-   body mass index
-   fasting glucose as a continuous outcome
-   metabolic syndrome as a binary outcome
-   annual follow-up visits as a count outcome

```{r, echo=FALSE}
set.seed(123)

N_pop <- 10000

population <- data.frame(
  id = 1:N_pop,
  age = rnorm(N_pop, mean = 60, sd = 10),
  bmi = rnorm(N_pop, mean = 28, sd = 5)
)

population$glucose <- rnorm(
  N_pop,
  mean = 100 + 0.8 * population$bmi,
  sd = 15
)

population$metabolic_syndrome <- rbinom(
  N_pop,
  size = 1,
  prob = plogis(-5 + 0.08 * population$bmi + 0.03 * population$age)
)

population$visits <- rpois(
  N_pop,
  lambda = exp(0.5 + 0.02 * population$age)
)

head(population)
```
:::

------------------------------------------------------------------------

## Population Mean of Fasting Glucose

::: {style="font-size:0.68em"}
-   **population mean fasting glucose**: average glucose level across all individuals in the full target population

-   entire simulated population is available $\rightarrow$ true value can be computed directly

```{r}
true_mean_glucose <- mean(population$glucose)
true_mean_glucose
```

-   true population mean fasting glucose for this simulated population: 122.26 mg/dL
-   represents a measure fasting glucose for every individual in the target population and represents the average across all of those individuals
:::

------------------------------------------------------------------------

### Population Proportion With Metabolic Syndrome

::: {style="font-size:0.68em"}
**metabolic syndrome**: clinical condition defined by the co-occurrence of multiple metabolic risk factors that increase the likelihood of cardiovascular disease and type 2 diabetes

Goal: find population proportion with metabolic syndrome - proportion of all individuals in the full population whose outcome value equals one

-   probability that a randomly selected patient from the target population has metabolic syndrome
:::

------------------------------------------------------------------------

### Population Proportion With Metabolic Syndrome

::: {style="font-size:0.68em"}
```{r}
true_prop_metabolic <- mean(population$metabolic_syndrome)
true_prop_metabolic
```

-   28.55% of individuals in the full population have metabolic syndrome
-   probability that a randomly selected individual from this population has metabolic syndrome is 0.2855
-   out of every 100 individuals in this population, about 29 would meet criteria for metabolic syndrome

Note: metabolic syndrome captures co-occurring factors like: - elevated glucose - abdominal obesity - dyslipidemia - hypertension
:::

------------------------------------------------------------------------

## Sampling From the Population

::: {style="font-size:0.68em"}
-   **sampling**: mechanism that generates the observed data used for estimation by sampling individuals from the broader population

-   sample below represents 150 individuals randomly drawn from the full population

Note: any quantity computed from the sample is only an estimate of the corresponding population quantity

```{r}
n <- 150

sample_data <- population[sample(1:N_pop, n), ]

head(sample_data)
```
:::

------------------------------------------------------------------------

## Estimating the Population Mean

::: {style="font-size:0.68em"}
-   sample mean fasting glucose: 121.81 mg/dL
-   average glucose level observed in this specific sample
-   true population mean: 122.2571 mg/dL
-   sample provides a reasonable approximation of the true population mean

```{r}
mean_glucose <- mean(sample_data$glucose)
mean_glucose
```
:::

------------------------------------------------------------------------

### Sampling Variability

::: {style="font-size:0.5em"}
`sampling variability` : describes how an estimate changes across repeated samples drawn from the same population

-   shows the range of values the estimator can take under repeated sampling

```{r, echo=FALSE}
#| fig-width: 4 
#| fig-height: 3
set.seed(123)

sample_means <- replicate(500, {
  s <- population[sample(1:N_pop, n), ]
  mean(s$glucose)
})

plot <- hist(
  sample_means,
  main = "Sampling Distribution of the \n Sample Mean for Glucose",
  xlab = "Sample Mean Glucose"
)
abline(v = true_mean_glucose, lwd = 2, lty = 2)

```
:::

------------------------------------------------------------------------

### Sampling Variability

::: {style="font-size:0.5em"}
**Plot Interpretation**

-   histogram shows sampling distribution of the sample mean
-   distribution shows that the estimate varies from sample to sample $\rightarrow$ cluster around the population mean glucose level
-   dashed vertical line represents the population mean
-   each bar represents a sample mean calculated from a different simulated study of glucose levels
-   every sample is drawn from the same underlying population $\rightarrow$ estimated mean is not identical across samples
-   distribution is approximately symmetric and centered around a value near 122
-   most of the sample means fall within a narrow range between 120 and 124

Note: a single observed estimate is one realization from a broader distribution of possible estimates

```{r, echo=FALSE}
#| fig-width: 4 
#| fig-height: 3
set.seed(123)

sample_means <- replicate(500, {
  s <- population[sample(1:N_pop, n), ]
  mean(s$glucose)
})

plot <- hist(
  sample_means,
  main = "Sampling Distribution of the \n Sample Mean for Glucose",
  xlab = "Sample Mean Glucose"
)
abline(v = true_mean_glucose, lwd = 2, lty = 2)

```
:::

------------------------------------------------------------------------

### Center and Spread of the Sampling Distribution

::: {style="font-size:0.68em"}
distribution summarizes two quantities:

-   mean of the sampling distribution
-   standard deviation of the sampling distribution

standard deviation: 1.21

-   describes how much the estimate varies from sample to sample
-   measures distance between an estimated mean and the true population mean across repeated studies

```{r}
c(
  mean_of_sample_means = mean(sample_means),
  sd_of_sample_means = sd(sample_means)
)
```
:::

------------------------------------------------------------------------

### Standard Error

::: {style="font-size:0.68em"}
**standard error**: metric that describes how much an estimate is expected to vary across repeated samples

`standard deviation (SD)`: describes how much individual observations vary

-   how different patients are from one another
-   how spread out the data are

`standard error (SE)` : describes how much the estimate varies

-   how much the sample mean would change across repeated studies
-   how precise the estimate is as a representation of the population
:::

------------------------------------------------------------------------

## Standard Error

::: {style="font-size:0.68em"}
```{r}
se_glucose <- sd(sample_data$glucose) / sqrt(nrow(sample_data))
se_glucose
```

-   standard error of the sample mean glucose: 1.207
    -   estimate of the amount that the sample mean would differ from the true population mean across repeated samples of the same size
    -   most sample means would fall within about ±1.21 units of the population mean
:::

------------------------------------------------------------------------

### Key Takeaways

::: {style="font-size:0.68em"}
-   estimation uses sample data to approximate unknown population quantities

-   research analysis is typically conducted on samples not entire populations

-   common quantities estimated include:

    -   mean biomarker levels
    -   probabilities of disease or response
    -   rates of clinical events
    -   associations between variables

-   estimates depend on the sample and vary as a result

-   different samples from the same population produce different estimates: `sampling variability`

-   estimates form a distribution centered near the population value across repeated samples, which is summarized by:

    -   standard deviation (SD) → variability in observed data
    -   standard error (SE) → variability of the estimate across samples
:::

------------------------------------------------------------------------
