---
title: "Statistical Estimation: Evaluating Bias, Variance, and Sampling Behavior"

author: "<br><br><span style='font-size:25px;'><strong>Lesley Chapman Hannah, Ph.D., M.S.</strong></span><br>College of Graduate Studies<br>Northeast Ohio Medical University"

format:
  revealjs:
    theme: simple
    slide-number: true
    transition: fade
    chalkboard: true
    smaller: false
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r setup, echo=FALSE}
library(tidyverse)
library(scales)

set.seed(104)

biomarker_df <- tibble(
  id = 1:180,
  treatment = rep(c("Placebo", "Low dose", "High dose"), each = 60),
  ifng = c(
    rnorm(60, mean = 11.5, sd = 4.2),
    rnorm(60, mean = 15.3, sd = 4.8),
    rnorm(60, mean = 19.8, sd = 5.1)
  ),
  response = c(
    rbinom(60, 1, 0.28),
    rbinom(60, 1, 0.45),
    rbinom(60, 1, 0.62)
  ),
  infections = c(
    rpois(60, lambda = 1.4),
    rpois(60, lambda = 1.1),
    rpois(60, lambda = 0.8)
  ),
  age = round(rnorm(180, mean = 61, sd = 10)),
  bmi = round(rnorm(180, mean = 29, sd = 4.5), 1)
)
```

### Overview

::: {style="font-size:0.5em"}
-   understanding how to identify ideal estimators that target a specific quantity and with less variation per sample

-   understand point estimates - sample-based value for a population parameter

-   characterize bias - description of whether an estimator is centered on the truth across repeated samples

-   discuss variance and standard error - descriptors of repeated-sampling
:::

------------------------------------------------------------------------

### Central Theme

::: {style="font-size:0.5em"}
**Recap**:

-   established that a sample statistic can be used to estimate an unknown population quantity
-   `estimator` : defines how the data are used to approximate the population quantity \[i.e.: mean, sample proportion, regression coefficient\]

**Current Aim**: Evaluate the quality of the estimator

Indicators of a good estimator:

1.  target the correct population quantity
2.  vary as little as possible across repeated samples

Estimator checks:

-   determine whether the estimator is centered on the target parameter

-   observe how widely the estimator varies across repeated samples

-   determine whether the design assumptions needed for that estimator are plausible

-   here the **sampling distribution** is a key tool for judging estimator quality
:::

------------------------------------------------------------------------

### Point Estimation

::: {style="font-size:0.5em"}
-   **point estimate** : single numerical value computed from sample data to approximate an unknown population parameter

-   interpretable quantities derived from data \[i.e.: average treatment effects, disease prevalence, or risk differences\] and serve as the basis for confidence intervals, hypothesis tests

Two examples include:

**Sample Mean**

$$
\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i
$$

and

**Sample Proportion**

$$
\hat{p} = \frac{x}{n}
$$

-   $\bar{x}$: sample mean $\rightarrow$ estimate of a population mean $\mu$
-   $x_i$: the value for the $i$th sampled observation
-   $n$: sample size
-   $\hat{p}$: sample proportion $\rightarrow$ estimate of a population proportion $p$
-   $x$: number of observations in the sample with the characteristic of interest
:::

------------------------------------------------------------------------

### Estimation in Biomedical Inference

::: {style="font-size:0.6em"}
-   estimation is used to quantify key quantities such as average biomarker level, disease prevalence, treatment response, or average processing time

-   biased estimator can systematically misstate the underlying biology or clinical burden

-   highly variable estimator can produce unstable results that differ substantially from one study sample to another
:::

------------------------------------------------------------------------

### Common Failures in Statistical Estimation

::: {style="font-size:0.6em"}
estimator can fail \[provide a poor estimation for the true population parameter\] in two ways:

-   exhibiting systematic deviation from the target [bias]

-   displaying excessive sampling variability \[high variance across repeated samples

-   estimator can fail because it is **biased**: estimator systematically misses the target

-   estimator lacks accuracy if it does not target the true parameter on average [bias]

-   estimator can be **too variable**: if estimator is centered correctly after multiple samples each individual sample-based estimates move around too much

-   estimator lacks precision if it is too variable across samples \[variance\]

The three panels therefore represent three distinct cases:

-   **biased estimator**: sampling distribution is shifted away from the true value \[systematic error\]
-   **unbiased imprecise estimator**: centered at the true value but with large spread \[high variability\]
-   **unbiased and precise estimator**: centered at the true value with small spread \[low variability\]

good estimation requires both correct centering \[low bias\] and tight concentration \[low variance\]
:::

------------------------------------------------------------------------

## Bias

::: {style="font-size:0.6em"}
`Bias` measures systematic deviation from the true parameter:

$$
\operatorname{Bias}(\hat{\theta}) = E(\hat{\theta}) - \theta
$$

Where:

-   $\hat{\theta}$: estimator

-   $E(\hat{\theta})$: expected value or average of the estimator across repeated samples

-   $\theta$: true population parameter being estimated

-   bias = zero : estimator is **unbiased**

-   bias = negative/positive value : average falls above or below the true parameter $\rightarrow$ estimator is **biased**
:::

------------------------------------------------------------------------

### Unbiased and Biased Estimators

::: {style="font-size:0.6em"}
-   **unbiased statistic** :
    -   sampling distribution centered at the true parameter
    -   over many repeated samples estimate does not overestimate or underestimate the target
-   **biased statistic** :
    -   estimator whose sampling distribution is centered away from the true parameter
        -   sampling distribution centered away from the target
:::

------------------------------------------------------------------------

### Precision and Variance

::: {style="font-size:0.6em"}
-   `precision` : describes how tightly an estimator clusters
-   estimator variance: expected squared distance between the estimator and the repeated-sample mean

Variance Formula

$$
\operatorname{Var}(\hat{\theta}) = E\left[(\hat{\theta} - E(\hat{\theta}))^2\right]
$$

-   $\operatorname{Var}(\hat{\theta})$: variance of the estimator

-   $\hat{\theta}$: random variable \[estimator\]

-   $E(\hat{\theta})$: center of the estimator across multiple samples

-   `smaller variance`: indicates that the estimator is more tightly concentrated around its expected value across repeated samples

-   `larger variance` : indicates that the estimator is more dispersed across repeated samples
:::

------------------------------------------------------------------------

### Accuracy Versus Precision

::: {style="font-size:0.6em"}
**accuracy** and **precision**: statistical properties that describe how an estimator’s sampling distribution relates to the true parameter value in terms of its center and spread

-   **accuracy [bias]** : metric to describe how close the center of the estimator’s sampling distribution is to the true parameter value

Example: accurate estimator is centered at the target in repeated sampling

-   **precision \[variability\]**: metric describing the spread of the estimator’s sampling distribution

Example: precise estimator has less sampling variability $\rightarrow$ repeated samples produce estimates that are more tightly clustered around the center of the distribution
:::

------------------------------------------------------------------------

## Research Question Background: Chronic Kidney Disease

::: {style="font-size:0.58em"}
-   chronic kidney disease (CKD): sustained reduction in kidney function and is a major contributor to morbidity and mortality at the population level
-   United States approximately 14–15% of adults are estimated to have CKD
-   kidney function is quantified using estimated glomerular filtration rate (eGFR) - continuous measure derived from serum creatinine and patient characteristics that approximates filtration capacity

estimated glomerular filtration rate \[eGFR\]

-   ≥ 90 → generally considered normal kidney function
-   60–89 → mildly reduced kidney function
-   \< 60 → clinically meaningful reduction, consistent with chronic kidney disease (CKD)
:::

------------------------------------------------------------------------

## Research Aim

::: {style="font-size:0.58em"}
Research goal(s) estimate two population quantities:

-   population mean eGFR
-   population proportion with CKD

Approach

1.  create a simulated biomedical population\
2.  generate simulated true population quantities
3.  draw study samples from that population\
4.  compute point estimates from each sample\
5.  examine repeated samples to study sampling distributions, bias, variance, precision
:::

------------------------------------------------------------------------

## Simulated Population Dataset

::: {style="font-size:0.50em"}
```{r, echo = FALSE}
library(tidyverse)
library(broom)

set.seed(123)

population <- tibble(
  id = 1:100000,
  age = pmin(pmax(rnorm(100000, mean = 59, sd = 12), 18), 90),
  systolic_bp = pmin(pmax(rnorm(100000, mean = 128, sd = 16), 90), 220),
  uacr = rgamma(100000, shape = 2.2, scale = 18),
  egfr = 118 - 0.72 * age - 0.11 * systolic_bp + rnorm(100000, mean = 0, sd = 10)
) |>
  mutate(
    egfr = pmax(egfr, 8),
    ckd = if_else(egfr < 60, 1, 0)
  )

head(population)
```

**Overview**

-   one row per individual in the population

-   **eGFR \[estimated glomerular filtration rate\]**:

    -   clinical measure of kidney function, expressed in mL/min/1.73 m²
    -   reflects how effectively the kidneys filter waste from the blood

-   **CKD \[chronic kidney disease\]** :

    -   long-term condition characterized by reduced kidney function
    -   binary indicator equal to 1 when eGFR is below 60 $\rightarrow$ indicator of the presence of disease

-   **age**: clinical predictor associated with physiologic decline in kidney function\

-   **systolic_bp**: blood pressure measure included because elevated blood pressure is clinically relevant to kidney disease risk\

-   **uacr**: urine albumin-to-creatinine ratio additional kidney function biomarker
:::

------------------------------------------------------------------------

## Population Quantities

::: {style="font-size:0.58em"}
Here we define the target quantities in the population:

-   population mean eGFR

-   population proportion with CKD

-   **population parameters** that the estimators are trying to recover

-   true average kidney function in the full population \[mean eGFR\]

```{r, echo=FALSE}
true_mean_egfr <- mean(population$egfr)

true_mean_egfr

```

-   true proportion of individuals classified as having CKD in this simulated population

```{r}
true_prop_ckd <- mean(population$ckd)

true_prop_ckd
```
:::

------------------------------------------------------------------------

### Determine Point Estimates

::: {style="font-size:0.58em"}
Following point estimates approximate the population quantity using only the information observed in the sample:

-   **sample mean eGFR** : point estimate because that reduces all observed eGFR values in the sample to one number that represents the population mean kidney function
-   **sample proportion with CKD** : point estimate because that reduces all observed disease indicators a single value that represents the population prevalence of CKD

Multiple individual observations are summarized in a single value representing the corresponding population parameter

```{r, echo=FALSE}
set.seed(321)

sample_data <- population |>
  slice_sample(n = 150)

sample_summary <- sample_data |>
  summarise(
    sample_mean_egfr = mean(egfr),
    sample_prop_ckd = mean(ckd)
  )

sample_summary
```

-   sample mean eGFR = 60.6 : average kidney function among the individuals who were included in the sample

-   sample proportion with CKD = 0.513 : 51.3% of individuals in this sample meet the CKD definition (eGFR \< 60)

-   estimates are close to the population values, the sample happened to produce estimates near true population parameters
:::

------------------------------------------------------------------------

### Sampling Distribution of the Sample Mean

::: {style="font-size:0.57em"}
-   **sampling distribution** : distribution of an estimator across many repeated samples of the same size

-   distribution shows both the center and the spread of the estimator

-   2,000 independent study samples of size 150 drawn randomly from the population - sample mean eGFR is calculated for each sample

-   `avg_estimate` : repeated-sample average of the estimator [bias]

-   `sd_of_estimates` : empirical standard deviation of the estimates\[sampling variability\]

```{r, echo=FALSE}
set.seed(100)

n_reps <- 2000
n_sample <- 150

sampling_results <- tibble(
  rep = 1:n_reps,
  sample_mean_egfr = map_dbl(rep, ~ {
    population |>
      slice_sample(n = n_sample) |>
      summarise(mean_egfr = mean(egfr)) |>
      pull(mean_egfr)
  })
)

out <- sampling_results |>
  summarise(
    avg_estimate = mean(sample_mean_egfr),
    sd_of_estimates = sd(sample_mean_egfr)
  )

out
```

-   average of the sample means across many repeated samples : 61.04
-   standard deviation of the sample means : 1.08
:::

------------------------------------------------------------------------

### Sampling Distribution of the Sample Mean

::: {style="font-size:0.50em"}
-   histogram shows the distribution of sample mean estimates across repeated samples

-   dashed line marks the true population mean eGFR

-   estimator is performing well $\rightarrow$ distribution is centered near the dashed line

-   narrower distribution indicates increased precision

```{r, echo=FALSE}
plot <- ggplot(sampling_results, aes(x = sample_mean_egfr)) +
  geom_histogram(bins = 35) +
  geom_vline(xintercept = true_mean_egfr, linetype = "dashed") +
  labs(
    title = "Sampling Distribution of the Sample Mean for eGFR",
    x = "Sample Mean eGFR",
    y = "Number of Repeated Samples"
  ) +
  theme_bw()

plot
```
:::

------------------------------------------------------------------------

## Unbiased and Biased Estimators

::: {style="font-size:0.55em"}
Goal: compare the following two estimators of mean eGFR:

-   estimator based on a simple random sample from the full population\
-   biased estimator based on a truncated sampling frame that excludes individuals with lower eGFR values

```{r, echo=FALSE}
set.seed(200)

bias_demo <- tibble(
  rep = 1:n_reps,
  unbiased_mean = map_dbl(rep, ~ {
    population |>
      slice_sample(n = n_sample) |>
      summarise(val = mean(egfr)) |>
      pull(val)
  }),
  biased_mean = map_dbl(rep, ~ {
    population |>
      filter(egfr >= 45) |>
      slice_sample(n = n_sample) |>
      summarise(val = mean(egfr)) |>
      pull(val)
  })
)

bias_results <- tibble(
  estimator = c("Unbiased sample mean", "Biased sample mean"),
  repeated_sample_mean = c(mean(bias_demo$unbiased_mean), mean(bias_demo$biased_mean)),
  bias = c(
    mean(bias_demo$unbiased_mean) - true_mean_egfr,
    mean(bias_demo$biased_mean) - true_mean_egfr
  )
)

bias_results
```

**Interpretation**

-   unbiased sample mean 61.4 - average of the sample mean estimates across many repeated random samples from the full population with 0.0163 bias

-   low bias indicates that on average the estimator recovers the true population mean with essentially no systematic error

-   64.2 : average of the sample mean estimates across repeated samples drawn from the truncated population \[i.e.: where lower eGFR individuals are excluded\]

-   2.79 : bias - estimator overestimates the true population mean by about 2.79 units on average
:::

------------------------------------------------------------------------

### Unbiased and Biased Estimators


```{r, echo=FALSE}

#| fig-width: 1
#| fig-height: 2


bias_demo_long <- bias_demo |>
  pivot_longer(
    cols = c(unbiased_mean, biased_mean),
    names_to = "estimator",
    values_to = "estimate"
  )

plot <- ggplot(bias_demo_long, aes(x = estimate)) +
  geom_histogram(bins = 30) +
  geom_vline(xintercept = true_mean_egfr, linetype = "dashed") +
  facet_wrap(~ estimator, scales = "free_y") +
  labs(
    title = "Biased and Unbiased Estimators of Mean eGFR",
    x = "Estimated Mean eGFR",
    y = "Number of Repeated Samples"
  ) +
  theme_bw()
plot

```

::: {style="font-size:0.50em"}

**Interpretation**

-   unbiased estimator : centered near the true mean eGFR
-   biased estimator : shifted upward away from the target because lower eGFR individuals were excluded before sampling

:::

------------------------------------------------------------------------

### Precision and Sample Size

::: {style="font-size:0.55em"}
-   compare sample estimator across different sample sizes
-   table shows the same estimator \[sample mean eGFR—behaves across repeated samples\] as the sample size increases

Note: target population mean changes very little while the precision of the estimator changes

-   `n` : sample size used for each draw
-   `mean_estimate` : average of the sample means across many repeated samples
-   `variance_estimate` and `sd_estimate` : quantify how much the sample mean varies from sample to sample

```{r, echo=FALSE}
simulate_means <- function(n_sample, n_reps = 2000) {
  tibble(
    rep = 1:n_reps,
    sample_mean = map_dbl(rep, ~ {
      population |>
        slice_sample(n = n_sample) |>
        summarise(val = mean(egfr)) |>
        pull(val)
    }),
    n = n_sample
  )
}

means_n30  <- simulate_means(30)
means_n150 <- simulate_means(150)
means_n500 <- simulate_means(500)

precision_demo <- bind_rows(means_n30, means_n150, means_n500)

precision_summary <- precision_demo |>
  group_by(n) |>
  summarise(
    mean_estimate = mean(sample_mean),
    variance_estimate = var(sample_mean),
    sd_estimate = sd(sample_mean)
  )

precision_summary
```

**Interpretation**

-   mean estimate remains stable (61.5 \[n=30\], 61.4 \[n=150\] and n=500)
-   indicates the estimator stays centered near the population mean
-   as sample size increases $\rightarrow$ variance decreases from 5.75 to 0.350 and the standard deviation decreases from 2.40 to 0.592
-   estimates become more precise across repeated samples
:::

------------------------------------------------------------------------

### Summary

::: {style="font-size:0.50em"}
-   analysis examples discussed began defining a population quantity \[parameter that cannot be directly observed
-   point estimate is a single numerical value computed from sample data to approximate that population parameter

Common estimators include:

-   sample mean → estimates population mean
-   sample proportion → estimates population prevalence

Sampling distribution describes how an estimator behaves across repeated samples

-   center → reflects bias (accuracy)
-   spread → reflects variance (precision)
:::

------------------------------------------------------------------------

### Summary

::: {style="font-size:0.50em"}
-   useful estimators generally have two properties:

    -   low bias \[accuracy\]

        -   estimator is centered near the true population parameter
        -   example: repeated mean ≈ 61.4 with negligible bias

    -   low variance \[precision\]

        -   estimates vary minimally across repeated samples

-   bias can arise from systematic issues in data collection or sampling design

    -   precision improves with larger sample size
:::

------------------------------------------------------------------------
